---
title: "Ex-02.1-pefr"
subtitle: ""
author:
 - name: Paul Johnson
   affiliation: CRMDA
   email: pauljohn@ku.edu
abstract: >
    RHS anorexia exercise
checked_by: "First Last"
Note to Authors: please_dont_change_the_next 4 lines!
date: "`r format(Sys.time(), '%b. %e %Y')`"
output:
  crmda::crmda_html_document:
    toc: true
    toc_depth: 2
    highlight: haddock
logoleft: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/jayhawk.png
logoright: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/CRMDAlogo-vert.png
---

# Data import
```{r import}
if (!file.exists("pefr.dta12")){
    download.file("http://www.stata-press.com/data/mlmus3/pefr.dta",
                  destfile = "pefr.dta12")
}
library(foreign)
pefr <- read.dta("pefr.dta12")
```

```{r recode}
rockchalk::summarize(pefr)
pefr$id.orig <- pefr$id
pefr$id <- as.factor(pefr$id)
```



## Visualize the imported data
```{r}
plot(wp1 ~ id.orig, data = pefr, main = "Wright Peak Flow for all Subjects")
points(wp2 ~ id, data = pefr, col = "red")
abline(h = (mean(pefr$wp1) + mean(pefr$wp2)) / 2)
legend("bottom", legend = c("pre", "post"), col = c("black", "red"),
       pch = c(1,1), horiz = TRUE)
```

# 2.1.1a: Reshape

Intention is to create "stacked" versions of "wm" and "wp"
variables. Try R's built-in reshape, this is an easy case,
a good one to explore the reshape function.

```{r}
head(pefr)
pefrlong <- reshape(pefr, direction="long",
                    varying = list(c("wp1", "wp2"), c("wm1", "wm2")),
                    v.names = c("wp", "wm"),
                    timevar = "occasion")
head(pefrlong, 10)
```

# 2.1.1b: mixed model with ML

Set `REML = FALSE` because we want maximum likelihood estimates
of the variance component.

```{r lmer10}
library(lme4)
m1 <- lmer(wp ~ (1|id), data = pefrlong, REML = FALSE)
summary(m1)
```

<!-- ```{r lmer20} -->
<!-- library(lmerTest) -->
<!-- m1b <- lmer(wp ~ (1|id), data = pefrlong, REML = FALSE) -->
<!-- summary(m1b) -->
<!-- ``` -->

```{r confint}
confint(m1)
```

# 2.1.2 Compare estimates for the wm and wp variables

First I'll fit the same mixed model with the `wm` outcome
variable. We need this for comparison.

```{r mw10}
m2 <- lmer(wm ~ (1|id), data = pefrlong, REML = FALSE)
summary(m2)
```

Lets compare the models side-by-side:

```{r outreg10, results = "hide"}
library(rockchalk)
or <- outreg(list("Wright Peak flow" = m1, "Mini Wright meter" = m2))
```

```{r outreg20, results = "asis"}
or <- gsub("[ ]{2,}", " ", or)
cat(or)
```

I did not understand this question in the RHS book, but the answer in
the teacher's manual points our attention to the estimate of the
variance of the residual in the fitted models.  They observe that the
residual error in the Wright peak-flow model (the one using `wp`) is
15.3, while the residual standard error in the other is 19.9.

The "reliability" of the measurement is the similarity of results when
applied to an equivalent case.  Here, we assume that the medical
condition of the patients did not change between occasion 1 and
occasion 2, so the observed fluctuation in their score is due, by
definition, to measurement error.


# 2.1.3: estimates for predictions of random effects

## Retrieve predictions of the random effects from m1

In the usual GLS interpretation of the mixed model, these are the 
empirical bayes est. of random intercept (RHS, p. 109-113). They
were often called "blups", but now it is recommended we refer to
them as posterior modes.

```{r blups10}
m1.ranef <- ranef(m1, condVar = TRUE)
m1.ranef
```
That demonstrates the predicted values of the "group-level"
intercepts, but here a "group" is one patient, observed twice.

There is a very handy way to visualize these random 
effects built into lme4.  The following is a "dotplot"
showing the random effects for each `id`. This includes
"error bars", which use information we will describe next.

```{r dotplot10}
library(lattice)
dotplot(m1.ranef)
```

The dotplot includes error bars to indicate uncertainty, a
95% confidence interval if you care to call it that. It is
created by a simple calculation, $ranef +/- 1.96\cdot std.err$.
Where do the standard errors come from?

The information about uncertainty associated with each posterior mode is tucked
away inside the object `m1.ranef`.  Here's how I know: I ask
for the attributes of the object:

```{r blups11a}
attributes(m1.ranef[["id"]])
```

If I want to create standard errors, I must extract those variance
values, and convert them to standard errors by taking the square root
of each one. Here one way. (See RHS, comparative standard error,
p. 113-114).

```{r blups11b}
m1.ranef.se <- sqrt( attr(m1.ranef[["id"]], "postVar")[1,1, ])
## Not easy. Get the 3-d array from attr(m1.ranef[["id"]], "postVar")
## Then snip off the part we need with [1, 1, ]
## Then take square root
```

## 2.1.3 Compare those random effects with the Mini Wright values


```{r mw15}
m2 <- lmer(wm ~ (1|id), data = pefrlong, REML = FALSE)
summary(m2)
m2.ranef <- ranef(m2, condVar = TRUE)
m2.ranef.se <- sqrt(attr(m2.ranef[["id"]], "postVar")[1,1, ])
```

```{r mw20}
## Not too easy to get the random effects as vectors, by themselves
y <- m1.ranef[["id"]]$`(Intercept)`
x <- m2.ranef[["id"]]$`(Intercept)`
plot(y ~ x,
     ylim = c(-300, 200), xlab = "Mini Wright meter predictions",
     ylab = "Wright peak flow predictions")
myx <- lm(y ~ x)
abline(myx, lty = 2)
abline(a = 0, b = 1, lty = 3)
legend("bottom", legend = c("least squares line", "y=x line"),
       lty = c(2, 3), horiz=TRUE)
```


# 2.1.4 Compare standard errors

The wm method:

```{r}
m2.ranef.se
```

```{r}
m1.ranef.se
```

Wait. Why are all of those the same number? This does not usually
happen when there are predictors which differentiate the cases and
there are different numbers of observations for each group. Here, we
have exactly 2 observations for each group (person), and the only
predictor is the `id` and thus the uncertainty for each random effect
is the same.
