---
title: "Ex-02.4-neighborhood"
subtitle: ""
author:
 - name: Paul Johnson
   affiliation: CRMDA
   email: pauljohn@ku.edu
abstract: >
    RHS neighborhood exercise
checked_by: "First Last"
Note to Authors: please_dont_change_the_next 4 lines!
date: "`r format(Sys.time(), '%b. %e %Y')`"
output:
  crmda::crmda_html_document:
    toc: true
    toc_depth: 2
    highlight: haddock
logoleft: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/logoleft.png
logoright: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/logo-vert.png
---

# Data import

```{r}
dsn <- "neighborhood"
if (!file.exists(paste0(dsn, ".dta12"))) {
    download.file(paste0("http://www.stata-press.com/data/mlmus3/", dsn, ".dta"),
                  destfile = paste0(dsn, ".dta12"))
}
library(foreign)
neighborhood <- read.dta(paste0(dsn, ".dta12"))
```

# Summarize
```{r}
head(neighborhood)

rockchalk::summarize(neighborhood)

str(neighborhood)
```


# Recode to factors

## The dummy variables
```{r}
makedummy <- function(x){
    if (is.integer(x)){
       xnew <-  as.factor(ifelse(x == 1, TRUE, FALSE))
       return(xnew)
    } else {
        MESSG <- "We refuse to make a numeric into a dummy"
        warning(MESSG)
        return(x)
    }
}

neighborhood$dadunemp <- makedummy(neighborhood$dadunemp)
neighborhood$daded <- makedummy(neighborhood$daded)
```

In a real project, I would not do that verbose method. I'd make
a vector of names and call in a  for loop like this.

```{r}
varstofix <- c("momed", "male")
for (i in varstofix) {
    neighborhood[ , i] <- makedummy(neighborhood[ , i])
}
str(neighborhood)
```


## The integers that will be used for classification variables

```{r}
varstofix <- c("neighid", "schid")
for (i in varstofix) {
    neighborhood[ , paste0(i, "f")] <-as.factor(neighborhood[ , i])
}
str(neighborhood)
```



# Random Effects models

## Model 1: Nested random effects
```{r}
library(lme4)

m1 <- lmer(attain ~ (1|schid), REML=FALSE, data = neighborhood)
summary(m1)



```

## The ICC
The icc can be calcuated after gazing at the output (seems dumb)

```{r}
0.08874 / (0.08874 + 0.93441)
```
Think of that as the correlation of students within a school. RHS
answer sheet says "In other words, 9% of the variance in attainment
is between schools."


## Model 2: Nested in neighborhood
```{r}
m2 <- lmer(attain ~ (1|neighid), REML=FALSE, data = neighborhood)
summary(m2)
```

# ML versus REML

```{r}
m1ml <- lmer(attain ~ (1|schid), data = neighborhood)
summary(m1ml)
```

```{r}
m2ml <- lmer(attain ~ (1|neighid), data = neighborhood)
summary(m2ml)
```


## Practice making adequate table



```{r}
or1 <- rockchalk::outreg(list("REML: m1" = m1, "ML: m1" = m1ml), type = "html")
or1 <- gsub("[ ]{2,}", " ", or1)
or2 <- rockchalk::outreg(list("REML: m2" = m2, "ML: m2" = m2ml), type = "html")
or2 <- gsub("[ ]{2,}", " ", or2)
```

Here is a table for the regression with school as a random effect:
```{r, results="asis"}
cat(or1)
```

Here is a table when the neighborhood is treated as a random effect:

```{r, results="asis"}
cat(or2)
```

