---
title: "Ex-01.1-HSB"
subtitle: ""
author:
 - name: Paul Johnson
   affiliation: CRMDA
   email: pauljohn@ku.edu
abstract: >
    RHS HSB exercise
checked_by: "First Last"
Note to Authors: please_dont_change_the_next 4 lines!
date: "`r format(Sys.time(), '%b. %e %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    highlight: haddock
---

# Data import
```{r import}
if (!file.exists("hsb.rds")){
    download.file("http://pj.freefaculty.org/guides/stat/DataSets/HSB/hsb.rds",
                  destfile = "hsb2.rds")
}
    hsb <- readRDS("hsb.rds")
    ## Or do it the hard way
    ## see: http://pj.freefaculty.org/guides/stat/DataSets/HSB/
    ## library(foreign)
    ## hsb.url <- "http://www.stata-press.com/data/mlmus3/hsb.dta"
    ## hsb <- read.dta(hsb.url)
    ## saveRDS(hsb, "hsb.rds")
```

# Descriptive Statistics
	
## Quick overview

```{r}
library(rockchalk)
rockchalk::summarize(hsb)
```

## Recodes needed

All of these variables present themselves as numeric variables, but
some are clearly categorical. Female is 0 or 1. The R tradition
would not have us treat that as a number, but rather it should be a
factor variable. Same with minority. Lets make factor versions of
them. So we need to make them into factor variables. 

I'll take some liberty here and rename in a way that is, in my
opinion, more correct. The variable is "gender" (not female), which
 takes values Male and Female.

```{r}
hsb$gender <- factor(hsb$female, levels= c(0, 1), labels = c("Male", "Female"))
hsb$ethnicity <- factor(hsb$minority, levels = c(0, 1), labels = c("White", "Non-white"))
hsb$sector <- factor(hsb$sector, levels = c(0, 1), labels = c("Public", "Catholic"))
```

Double-check the recoding

```{r}
with(hsb, table(gender, female))
with(hsb, table(ethnicity, minority))
```

If you are comfortable with those recodes, you can move on.  I'm going
 to remove the original variables, just to show I can.

```{r}
hsb$female <- NULL
hsb$minority <- NULL
```

If we were going to do any "multi-level" regressions, we would need
the school variable as a factor, not as an integer. I'm going to make
that one into a factor called schoolidf

```{r}
hsb$schoolidf <- as.factor(hsb$schoolid)
```

Plainly, there are other variables in the collection that need the same
treatment, but I'm not bothering with them now. 

```{r}
rockchalk::summarize(hsb)
str(hsb)
```

## Describe Math Achievement with histograms

```{r}
rockchalk::summarize(hsb$mathach, probs = c(0, .25, .50, .75, 1.0))
```

```{r}
with(hsb, hist(mathach))
```


# 1.1.2 School and variable subsets


I'll keep some additional variables
```{r}
keepers <- c("schoolid", "schoolidf", "mathach", "ses", "ethnicity", "gender")
hsb2 <- hsb[hsb$schoolid <= 1317, keepers]
```

```{r}
summarize(hsb2)
```

```{r}
hsbmeans <- aggregate(hsb2[ , c("mathach", "ses")], by = list(schoolidf = hsb2$schoolidf), mean, na.rm = TRUE)
hsbsds <- aggregate(hsb2[ , c("mathach", "ses")], by = list(schoolidf = hsb2$schoolidf), sd, na.rm = TRUE)
merge(hsbmeans, hsbsds, by = "schoolidf", suffixes = c("mean", "sd"))                  
```


# Plots

## 1.1.3 histogram

```{r}
hist(hsb2$mathach, main = "Math achievement", xlab = "Math")
```

## boxplot


```{r}
plot(mathach ~ gender, data = hsb2)
```

Note: plot was a shortcut

```{r}
boxplot(mathach ~ gender, data = hsb2)
```

Beautify with labels:

```{r}
bp1 <- boxplot(mathach ~ gender, data = hsb2)

text(1, 14.3, "Median of Male")
text(2, 12.1, "Median of Female")
```


## 1.1.4 Show a scatterplot of 2 "numeric" variables

I'll adjust range to look like the stata output
```{r}
plot(mathach ~ ses, data = hsb2, xlim = c(-2, 2),
     ylim = c(-10, 30))
```

## Trellis plot with R lattice

```{r}
library(lattice)
xyplot(mathach ~ ses | schoolidf, data = hsb2,
        xlim = c(-2, 2),
     ylim = c(-10, 30))
```

The ggplot2 variant of same
```{r}
library(ggplot2)
theme_set(theme_bw())
ggplot(data = hsb2, aes(x=ses, y=mathach)) +
  geom_point()
```

```{r}
library(ggplot2)
theme_set(theme_bw())
ggplot(data = hsb2, aes(x=ses, y=mathach)) +
  geom_point() +
  facet_wrap(~ schoolid)
```


## Make list of subsets with split and plot them. 
```{r}
hsbsplit <- split(hsb2, hsb2$gender)

par(mfrow = c(2,1))
lapply(hsbsplit, function(subdat){
    with(subdat, hist(mathach, prob = TRUE,
                      main = paste("Math Achievement:",
                                   unique(subdat$gender))))})
```

```{r}
lapply(hsbsplit, function(subdat) {
    with(subdat, hist(ses, prob = TRUE, xlim = c(-4, 3),
                      main = paste0("SES:", unique(subdat$gender))))})
```

## 1.1.5 Regression

```{r}
m0 <- lm(mathach ~ ses, data = hsb2)
```

```{r, results="hide"}
m0.or <- outreg(list("Linear Model" = m0), type = "html", browse = FALSE)
```

```{r, results = "asis"}
m0.or <- gsub("[ ]{2,}", " ", m0.or)
cat(m0.or)
```

```{r}
confint(m0)
```

## 1.1.6 Predict

```{r}
hsb2$yhat <- predict(m0)
head(hsb2)
```

```{r}
plotSlopes(m0, plotx = "ses")
```

```{r}
m0.pom <- predictOMatic(m0, interval="confidence")
m0.pom$ses$mathach <- m0.pom$ses$fit
m0.pom
```

Maybe you like a confidence interval as well

```{r}
plotSlopes(m0, plotx = "ses", interval = "confidence", xlim = c(-2, 2),
           ylim = c(-10, 30))
```


What best way to get plot for each school.

My plotSlopes approach is not what they want

```{r}
m1 <- lm(mathach ~ ses * schoolidf, data = hsb2)
plotSlopes(m1, plotx = "ses", modx = "schoolidf", xlim = c(-2, 2),
           ylim = c(-10, 30))
```

```{r}
xyplot(mathach ~ ses | schoolidf, data = hsb2, type = c("p", "r"))
```

```{r}
library(ggplot2)
theme_set(theme_bw())
ggplot(data = hsb2, aes(x=ses, y=mathach)) + geom_point(alpha=0.5) +
    geom_line(data = m0.pom$ses) +
    geom_ribbon(data = m0.pom$ses, aes(ymin = lwr, ymax = upr), fill = "pink", alpha = 0.5)
```

```{r}
ggplot(data = hsb2, aes(x=ses, y=mathach)) + geom_point(alpha=0.5) +
    geom_line(data = m0.pom$ses) +
    geom_ribbon(data = m0.pom$ses, aes(ymin = lwr, ymax = upr), fill = "pink", alpha = 0.5) +
    geom_smooth(method="lm", fill = NA) + 
    facet_wrap(~ schoolidf)
```

## 1.1.8 Regression with school as a factor


I already ran this one in previous, but I used interaction, 
not an additive model. So here it is, additive!

```{r}
m2 <- lm(mathach ~ ses + schoolidf, data = hsb2)
summary(m2)
```

## Test the school coefficients, jointly

```{r}
anova(m0, m2)
```

## 1.1.9 Lets see if there is an interaction between ses and school

```{r}
m3 <- lm(mathach ~ ses * schoolidf, data = hsb2)
summary(m3)
```




## Lets see if there is an interaction between ses and gender

```{r}
m1 <- lm(mathach ~ ses * gender, data = hsb2)
summary(m1)
```

```{r}
m2 <-  lm(mathach ~ ses + gender, data = hsb2)
summary(m2)
```

```{r}
library(rockchalk)
plotSlopes(m1, plotx = "ses", modx = "gender")
```
```{r}
plotSlopes(m1, plotx = "ses", modx = "gender", col = c("orange", "blue"))
```

```{r}
plotSlopes(m1, plotx = "ses", modx = "gender",
           col = gray.colors(2, 0.1, 0.5),
           main = "50 - 48 Shades of Gray")
```

```{r}
mycol <- c("black", "gray40")
mypch <- c(1, 8) 
plotSlopes(m1, plotx = "ses", modx = "gender", col = mycol, pch = mypch[hsb2$gender])
```

```{r}
plotSlopes(m2, plotx = "ses", modx = "gender", col = mycol, pch = mypch[hsb2$gender] )
```

## Compare the p-value from the anova test  

```{r}
anova(m1, m2)
```
