---
title: "Ex-01.3-smoking"
subtitle: ""
author:
 - name: Paul Johnson
   affiliation: CRMDA
   email: pauljohn@ku.edu
abstract: >
    RHS smoking exercise
checked_by: "First Last"
Note to Authors: please_dont_change_the_next 4 lines!
date: "`r format(Sys.time(), '%b. %e %Y')`"
output:
  crmda::crmda_html_document:
    toc: true
    toc_depth: 2
    highlight: haddock
logoleft: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/jayhawk.png
logoright: /home/pauljohn/R/x86_64-pc-linux-gnu-library/3.4/crmda/theme/CRMDAlogo-vert.png
---

# Data import
```{r import}
if (!file.exists("smoking.dta12")){
    dsname <- "http://www.stata-press.com/data/mlmus3/smoking.dta"
    download.file(dsname,
                  destfile = "smoking.dta12")
}
library(foreign)
dat <- read.dta("smoking.dta12")
rockchalk::summarize(dat)
dim(dat)
dat <- dat[dat$idx==1, ]
dim(dat)
```

# 1.3.2 Recode

```{r recode10}
dat$education.num <- ifelse(dat$hsgrad == 1, 1,
                        ifelse(dat$somecoll == 1, 2,
                        ifelse(dat$collgrad == 1, 3, 0)))
dat$education <- factor(dat$education.num, levels = 0:3,
                        labels = c("none", "hs", "somecoll", "collgrad"))
table(dat$education, dat$education.num)
```

# 1.3.3 Summary stats

I thought about going at this by making a newdata structure, but in
end was more trouble than it was worth.

```{r}
m1 <- lm(birwt ~ education + male + black, data = dat)
summary(m1)
dat2 <- rockchalk::newdata(m1, list("education" = unique(dat$education), "male" = unique(dat$male), "black" = unique(dat$black)))
```

So I fall back on old split/lapply approach. I don't have patience to
re-align the output to be in same style, but could do with ftabs or
table package.  Would be good assignment for GRA.

```{r}
splits <- split(dat, f = list(dat$male, dat$black, dat$smoke, dat$education), drop = TRUE)
reslts <- lapply(splits, function(ds){
    data.frame("male" = unique(ds$male), "black" = unique(ds$black),
               "smoke" = unique(ds$smoke), "education" = unique(ds$education),
               "N" = NROW(ds), 
               "mean" = mean(ds$birwt, na.rm=TRUE), "sd" = sd(ds$birwt, na.rm=TRUE))})
sumrytable <- do.call("rbind", reslts)
rownames(sumrytable) <- NULL
sumrytable
```

# 1.3.4 boxplots. 

For some reason, I'm not getting this split quite right

```{r boxplot}
library(lattice)
bwplot(birwt ~  education + black | smoke + male, data = dat)
```


# 1.3.5 Regression

```{r reg10}
m1 <- lm(birwt ~ smoke, data = dat)
summary(m1)
```

# 1.3.6 Regression #2

```{r reg20}
m2 <- lm(birwt ~ smoke + mage + male + black + education, data = dat)
summary(m2)
```

```{r reg30}
rockchalk::predictOMatic(m2, "margins", interval = "confidence")
```


```{r reg40}
rockchalk::predictOMatic(m2, "auto", interval = "confidence")
```
